---
title: "Carnivore diets"
author: "Peter-JL"
date: "2022-08-15"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(widyr)
library(tidytext)
theme_set(theme_light())

#data from github: 
#https://github.com/osmiddleton/CarniDIET-Database
#https://onlinelibrary.wiley.com/doi/10.1111/geb.13296
carnivore_diets_raw <- read_csv("https://raw.githubusercontent.com/osmiddleton/CarniDIET-Database/master/Version%201.0/CarniDIET%201.0.csv")

#data cleaning
carnivore_diets <- carnivore_diets_raw %>% 
  rename(carnivore = commonNameCarni,
         prey = commonNamePrey) %>% 
  mutate(prey = ifelse(str_detect(prey, "Snowshoe"), "Snowshoe hare", prey)) #get rid of weird character, "Snowshoe hare\xe1"
#note: should filter for taxonRankPrey=="Species"
```


Can I just use common name? one to one correspondence between scientific and common name?
```{r}

carnivore_diets %>% 
  distinct(scientificNameCarni, carnivore) %>% 
  count(scientificNameCarni, sort=TRUE)

```

each scientific name corresponds to only one common name, will just use common name. Also there's 103 carnivore species in this data set.


Break down of carnivores, limit to  top 50

```{r}
carnivore_diets %>% 
  filter(taxonRankPrey=="Species",
         fct_lump(carnivore, 50)!="Other") %>% 
  count(carnivore, familyCarni, sort = TRUE) %>% 
  mutate(carnivore = fct_reorder(carnivore, n)) %>% 
  ggplot(aes(n, carnivore, colour = familyCarni)) +
  geom_point() +
  scale_x_log10() +
  labs(x="# of prey species consumed (log)",
       y="",
       title = "Carnivores with the most diverse diets")
```


Same data but different graph, organized by family of carnivore

```{r}
carnivore_diets %>% 
  filter(taxonRankPrey=="Species",
         fct_lump(carnivore, 50)!="Other") %>% 
  count(carnivore, familyCarni, sort = TRUE) %>% 
  mutate(carnivore = reorder_within(carnivore, n, familyCarni),
         familyCarni=fct_lump(familyCarni, 3),
         familyCarni = fct_reorder(familyCarni, -n, sum)) %>% 
  ggplot(aes(n, carnivore, colour = familyCarni)) +
  geom_point() +
  #scale_x_log10() +
  scale_y_reordered() +
  facet_wrap(~familyCarni, scales = "free") +
  labs(x="# of prey species consumed (log)",
       y="",
       title = "Carnivores with the most diverse diets")
```


How many prey do carnivores eat, distribution?

```{r}
carnivore_diets %>% 
  filter(taxonRankPrey=="Species") %>% 
  count(carnivore) %>% 
  ggplot(aes(n)) +
  geom_histogram() +
  scale_x_log10() +
  labs(x="# distinct species eaten by carnivores (log)")
```

one carnivore eats over 4,000 distinct types of prey?

```{r}
carnivore_diets %>% 
  filter(taxonRankPrey=="Species") %>% 
  count(carnivore, sort=TRUE)
```

The Red fox eats 4,335 different types of prey.


Break down of prey, limit to top 50

```{r}
carnivore_diets %>% 
  filter(taxonRankPrey=="Species", #limit to species
         fct_lump(prey, 50)!="Other") %>% 
  count(prey, sort=TRUE) %>% 
  mutate(prey = fct_reorder(prey, n)) %>% 
  ggplot(aes(n, prey)) +
  geom_point() +
  #scale_x_log10() +
  labs(x="# of carnivores who consume this prey",
       title = "Most common prey species")
```

There seems to be carnivores who are also prey

```{r}
carnivore_diets %>% 
  pivot_longer(c(carnivore,prey), names_to="type", values_to="species") %>% 
  distinct(type, species) %>% 
  count(species, sort=TRUE) %>% 
  filter(n==2)
```

There are 55 species who are both carnivores and prey, e.g., American badger, American marten... and Brown Bear?


Maybe Prey species is too vague, maybe familyPrey is a better variable to use (more consistency in naming, as opposed to 0.1-0.3kg rodent which might also be named 0.3-0.5 kg rodent).

```{r}
carnivore_diets %>% 
  filter(fct_lump(familyPrey, 50)!="Other") %>% 
  distinct(carnivore, familyPrey, .keep_all = TRUE) %>% 
  count(familyPrey, sort=TRUE) %>% 
  mutate(familyPrey = fct_reorder(familyPrey, n)) %>% 
  ggplot(aes(n, familyPrey)) +
  geom_point() +
  scale_x_log10() +
  labs(x="# of carnivores who consume this family of prey (log)",
       title = "Most common prey families")
```


Which prey are most often eaten by the same predator?

```{r}
carnivore_diets %>% 
  filter(taxonRankPrey=="Species") %>% 
  count(carnivore, prey, sort=TRUE) %>% 
  add_count(prey) %>% 
  filter(nn>4) %>% 
  pairwise_cor(prey, carnivore, n, sort=TRUE)
```



How about family of prey?

```{r}
carnivore_diets %>% 
  filter(!is.na(familyPrey)) %>% 
  count(carnivore, familyPrey, sort=TRUE) %>% 
  filter(n>100) %>% 
  pairwise_cor(familyPrey, carnivore, n, sort=TRUE)
```



SVD, which prey are best able to distinguish predators?

```{r}
prey_svd <- carnivore_diets %>% 
  filter(taxonRankPrey=="Species", 
         !is.na(prey)) %>% 
  count(carnivore, prey, sort=TRUE) %>% 
  mutate(value=1) %>% 
  widely_svd(prey, carnivore, n)
prey_svd %>% 
  filter(between(dimension, 2, 5)) %>% 
  group_by(dimension) %>% 
  top_n(abs(value), n=10) %>% 
  ungroup() %>% 
  mutate(prey = reorder_within(prey, value, dimension)) %>% 
  ggplot(aes(value, prey, fill=value>0)) +
  geom_col() +
  scale_y_reordered() +
  facet_wrap(~dimension, scales = "free_y")
```

* 1st sv = small vs large animals
* 2nd sv = sambar & chital are indian deer, langur is indian monkey vs prey in the americas




What countries are there?

```{r}
carnivore_diets %>% 
  count(country, sort=TRUE)
```

which country has the most carnivore species?

```{r}
carnivore_diets %>% 
  distinct(carnivore, .keep_all=TRUE) %>% 
  count(country, sort=TRUE)
```

This might be a US-centric data set.







aggregate by carnivore, limit to carnivores with at least 5 distinct prey?

```{r}
carnivore_diets %>% 
  group_by(carnivore) %>% 
  summarize(ncountries = n_distinct(country, na.rm=TRUE),
            nprey = n_distinct(prey, na.rm=TRUE),
            lat = mean(decimalLatitude, na.rm = TRUE),
            long = mean(decimalLongitude, na.rm = TRUE), 
            pct_birds = mean(prey=="Birds", na.rm=TRUE),
            pct_fish = mean(prey=="Fish", na.rm=TRUE),
            pct_plants_seeds = mean(prey %in% c("Plant material", "Fruit/seeds"), na.rm=TRUE),
            pct_food_mammal = mean(foodType=="Mammal", na.rm=TRUE),
            pct_food_bird = mean(foodType=="Bird", na.rm=TRUE),
            pct_plants = mean(foodType=="Plant", na.rm=TRUE))

```


